---
title: "Feature preparation for climate variables"
author: 'Gabriel Munoz'
date: today
---

## Load libraries

```{r}
library(raster)
library(rasterVis)
library(gridExtra)
library(viridis)
library(grid)
library(tidyverse)
```

Make a function to compute the temporal slope between the change of temperature between time bins


```{r}
#' Calculate the temporal slope or residual variance for each grid cell in a raster stack
#'
#' This function computes either the temporal slope or the variance of the residuals 
#' from a linear regression for each grid cell in a given raster stack. The slope 
#' describes the trend over time for each cell, while the residual variance gives 
#' an indication of the variability around the trend.
#'
#' @param raster_stack A raster stack or brick where each layer represents a time point.
#' @param which A character string indicating whether to return the 'slope' (default) 
#' or the 'var' (residual variance) for each grid cell. The 'slope' indicates the rate 
#' of change over time, and 'var' reflects the variability of the model residuals.
#'
#' @return A raster object containing the computed slope or residual variance for each 
#' grid cell.
#'
#' @details The function extracts the raster values as a matrix, where each row corresponds 
#' to a spatial grid cell and each column to a time point. A linear regression is then 
#' fitted to the values for each cell, and either the slope or the variance of the 
#' residuals from this model is returned, depending on the argument provided.
#'
#' @note If all values for a grid cell are `NA`, the result for that cell will also be `NA`.
#'
#' @importFrom raster nlayers as.matrix values
#' @importFrom stats lm coef residuals sd
#'
#' @examples
#' # Example usage:
#' library(raster)
#' # Create a raster stack with random values (for demonstration)
#' r1 <- raster(matrix(runif(100), 10, 10))
#' r2 <- raster(matrix(runif(100), 10, 10))
#' r3 <- raster(matrix(runif(100), 10, 10))
#' raster_stack <- stack(r1, r2, r3)
#' 
#' # Calculate the temporal slope for each grid cell
#' slope_raster <- get_temporal_slope(raster_stack, which = 'slope')
#'
#' # Calculate the residual variance for each grid cell
#' variance_raster <- get_temporal_slope(raster_stack, which = 'var')
#'
#' @export
#' 
get_temporal_slope <- function(raster_stack, which = c('slope', 'var')){
  

  
  # Get the number of layers and create a time variable
  num_layers <- nlayers(raster_stack)
  time_points <- rev(1:num_layers)

  # Extract raster values as a matrix (each column represents a time point)
  values_matrix <- as.matrix(raster_stack)
  
  # Function to calculate the slope of the regression line for each grid cell
  calc_slope <- function(values, time_points, which = c('slope', 'var')) {
    if (all(is.na(values))) {
      return(NA)
    } else {
      lm_model <- lm(values ~ time_points)
      
      
      if (which == 'var') {
        return(sd(residuals(lm_model)))  # Variance of the model residuals
      } else {
        return(coef(lm_model)[2])  # Slope of the regression line
      }
    }
  }
  
  # Apply the function to each grid cell
  slope_values <- apply(values_matrix, 1, calc_slope, time_points, which)
  
  # Create a raster for the slope values
  slope_raster <- raster_stack[[1]]
  values(slope_raster) <- slope_values
  
  return(slope_raster)
  
}


```

## Making climate features from temperature spatiotemporal data 

```{r}

Temp_binned  <- readRDS( 'DATA/temperature_grid/Temp_binned.rds')

## set the adequate projection

# Lambert Azimuthal Equal-Area for Europe
# Define the Lambert Azimuthal Equal-Area projection centered between NA and Europe
crs_equal_area <- CRS("+proj=laea +lat_0=45 +lon_0=-30 +datum=WGS84 +units=m")

Temp_binned <-
  Temp_binned |>
  map(function(x) {

    crs(x) <- CRS("+proj=longlat +datum=WGS84 +no_defs")

    raster::projectRaster(x, crs = crs_equal_area)

  })


```


## Mean temperature 

```{r, eval = TRUE}
mean_temp <- sapply(Temp_binned, function(stack) {
  mean(stack, na.rm = TRUE)
})


```


```{r, echo = FALSE} 
saveRDS(mean_temp, 'DATA/temperature_grid/mean_temp.rds')

mean_temp <- readRDS('DATA/temperature_grid/mean_temp.rds')

library(terra)



```


```{r}
mean_temp_stack <- stack(mean_temp)
plot(mean_temp_stack)

saveRDS(mean_temp_stack, 'DATA/out/mean_temp_stack.rds')


```

## Temperature gradient


```{r}
temp_gradients <- lapply(mean_temp, function(stack) {
  terrain(stack, opt = "slope", 
          unit = "degrees", neighbors = 8)
})

temp_gradients <- stack(temp_gradients)

plot(temp_gradients)

saveRDS(temp_gradients, 'DATA/out/temp_gradients.rds')

```


## Temporal slope


```{r}
temporal_slope <- Temp_binned |> map(function(x) {
  get_temporal_slope(x, which = 'slope')})

temporal_slope <- stack(temporal_slope)

plot(temporal_slope)
saveRDS(temporal_slope, 'DATA/out/temporal_slope.rds')

```

## Temporal variance 

```{r}
temporal_variance <- Temp_binned |> map(function(x) 
  get_temporal_slope(x, which = 'var'))

temporal_variance <- stack(temporal_variance)

plot(temporal_variance)

saveRDS(temporal_variance, 'DATA/out/temporal_variance.rds')

```


# Temperature deltas 


## Shifts in mean temperature 

```{r}

# Number of layers
n_layers <- nlayers(mean_temp_stack)

# Compute differences using vectorized operations
diff_mean_temp <- mean_temp_stack[[2:n_layers]] - mean_temp_stack[[1:(n_layers - 1)]]

# Find the min and max values across all layers for unified scale
min_val <- min(diff_mean_temp, na.rm = TRUE)
max_val <- max(diff_mean_temp, na.rm = TRUE)

# Convert the raster brick to a data frame for ggplot2
brick_df <- as.data.frame(diff_mean_temp, xy = TRUE)
brick_df_melt <- reshape2::melt(brick_df, id.vars = c("x", "y"))

# Plot all layers with the same color scale
p<- ggplot(brick_df_melt, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  scale_fill_viridis_c(option = 'B') +  # Unified color scale
  theme_minimal() +
  labs(fill = "Legend") +
  # Use transition_states to animate the 'layer'
  transition_states(states = variable, transition_length = 2, state_length = 1) +
  ease_aes('linear')  # Optional: Makes the transitions smooth


# Animate the plot
anim <- animate(p, nframes = n_layers, fps = 2)  # Adjust nframes and fps for speed
```

## Shifts in spatial similarity 

```{r}
# Compute differences using vectorized operations
diff_temp_gradients <- temp_gradients[[2:n_layers]] - temp_gradients[[1:(n_layers - 1)]]


# Find the min and max values across all layers for unified scale
min_val <- min(diff_temp_gradients, na.rm = TRUE)
max_val <- max(diff_temp_gradients, na.rm = TRUE)

# Convert the raster brick to a data frame for ggplot2
brick_df <- as.data.frame(diff_temp_gradients*1000, xy = TRUE)
brick_df_melt <- reshape2::melt(brick_df, id.vars = c("x", "y"))

# Plot all layers with the same color scale
p <- ggplot(brick_df_melt, aes(x = x, y = y, fill = value)) +
  geom_raster() +  # Plot each layer separately
  scale_fill_viridis_c(option = 'B') +  # Unified color scale
  theme_minimal() +
  labs(fill = "Legend") +
  # Use transition_states to animate the 'layer'
  transition_states(states = variable, transition_length = 2, state_length = 1) +
  ease_aes('linear')  # Optional: Makes the transitions smooth


# Animate the plot
anim <- animate(p, nframes = n_layers, fps = 2)  
anim
```

## Shifts in climate change velocity (cliamte change acceleration)

```{r}
# Compute differences using vectorized operations
diff_temporal_slopes <- temporal_slope[[2:n_layers]] - temporal_slope[[1:(n_layers - 1)]]

names(diff_temporal_slopes) <- paste(names(temporal_slope)[-8],names(temporal_slope)[-1], sep = '_to_')

# Find the min and max values across all layers for unified scale
min_val <- min(diff_temporal_slopes, na.rm = TRUE)
max_val <- max(diff_temporal_slopes, na.rm = TRUE)

# Convert the raster brick to a data frame for ggplot2
brick_df <- as.data.frame(diff_temporal_slopes, xy = TRUE)
brick_df_melt <- reshape2::melt(brick_df, id.vars = c("x", "y"))



# Plot all layers with the same color scale
p <- ggplot(brick_df_melt, aes(x = x, y = y, fill = value)) +
  geom_raster() +  # Plot each layer separately
  scale_fill_viridis_c(option = 'B') +  # Unified color scale
  theme_minimal() +
  labs(fill = "Legend") +
  # Use transition_states to animate the 'layer'
  transition_states(states = variable, transition_length = 2, state_length = 1) +
  ease_aes('linear')  # Optional: Makes the transitions smooth


# Animate the plot
anim <- animate(p, nframes = n_layers, fps = 2)  
anim
```


## Shifts in the temporal instability 

```{r}
# Compute differences using vectorized operations
diff_temporal_variance <- temporal_variance[[2:n_layers]] - temporal_variance[[1:(n_layers - 1)]]

names(diff_temporal_variance) <- paste(names(temporal_variance)[-8],names(temporal_variance)[-1], sep = '_to_')

# Find the min and max values across all layers for unified scale
min_val <- min(diff_temporal_variance, na.rm = TRUE)
max_val <- max(diff_temporal_variance, na.rm = TRUE)

# Convert the raster brick to a data frame for ggplot2
brick_df <- as.data.frame(diff_temporal_variance, xy = TRUE)
brick_df_melt <- reshape2::melt(brick_df, id.vars = c("x", "y"))


# Plot all layers with the same color scale
p <- ggplot(brick_df_melt, aes(x = x, y = y, fill = value)) +
  geom_raster() +  # Plot each layer separately
  scale_fill_viridis_c(option = 'B') +  # Unified color scale
  theme_minimal() +
  labs(fill = "Legend") +
  # Use transition_states to animate the 'layer'
  transition_states(states = variable, transition_length = 2, state_length = 1) +
  ease_aes('linear')  # Optional: Makes the transitions smooth


# Animate the plot
anim <- animate(p, nframes = n_layers, fps = 2)  
anim
```



